A Multiresolution Framework for One Dimensional Neural Field Modelling

### 2011-02-27 Multiresolution_framework/ltx/manuscript.tex
edit to EM

%and  $\boldsymbol\lambda(\mathbf{s}-\mathbf{r})$ is the vector of the kernel scaling and wavelet functions and   $\boldsymbol\mu(\mathbf{s})$ is the vector of the field scaling and wavelet functions. 

@@

The  $m$th order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 
\begin{eqnarray}
N_{m}\left(s\right)&=&\left(N_{m-1}\ast N_{1}\right)\left(s\right)\nonumber \\
&=&\int_0^{1} N_{m-1}\left( s-r\right)dr \quad m>1 
\label{SplineConvolutionIntegral}
\end{eqnarray}
Here $\ast$ denotes the convolution operator and $N_1\left(s\right)$ is the characteristic function of the unit interval  $\left[ 0,1\right) $
\begin{equation}
N_{1}\left(s\right)=
\begin{cases}
1 & \text{if $ 0\le s<1$}, \\
0 & \mathrm{elsewhere}
\end{cases}
\end{equation}
In order to compute the  integrals (\ref{eq:ConstructionIntegral1}) and (\ref{eq:ConstructionIntegral2}) we need to be able to calculate the convolution and inner product of B-spline scaling functions. Equation (\ref{SplineConvolutionIntegral}) can be rewritten as $(m-1)$ times convolution of indicator function with itself 
\begin{equation}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right)
\end{equation}
then by using the associativity property of the convolution operator, it is clear that
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right)
\end{align}
A direct consequence of \eqref{eq:BsplineConvolution} is \cite{Chui1992} 
\begin{eqnarray}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle&=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
&=&N_{m+m'}\left(m'+l_{2}-l_{1}\right)
\label{eq:BsplineInnerProduct}
\end{eqnarray}
where $\left\langle  .,.\right\rangle $ denotes the inner product.

@@

\section{Estimation of the Multiresolution Spatio-temporal IDE Model}
A state-space representation of the IDE model allows the use of the Expectation Maximisation (EM) algorithm to infer both the kernel and the field from the electrophysiological data. The EM algorithm is a two part iterative algorithm that yields the maximum likelihood estimate of the connectivity kernel parameters and the posterior distribution of the field weights. In the M-step a lower bound $\mathcal{Q}\left(\boldsymbol\theta,\boldsymbol\theta'\right)$ of the complete-data log-likelihood  $p\left(\mathbf{X},\mathbf{Y} \mid \boldsymbol\theta \right)$, is maximised under the assumption of the system parameters being $\boldsymbol\theta'$ \cite{McLachlan1997}. In the E-step, given an estimate of $\boldsymbol\theta$, the Rauch Tung Striebel (RTS) smoother \cite{Gibsona2005} can be employed to determine the posterior state sequence. 

@@

where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The second terms in equations (\ref{eq:upsilon0}-\ref{eq:Upsilon}) can be computed before EM implementation which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}. $\Xi$-variables can be calculated using the smoother outputs, i.e., state estimates, $\hat{\mathbf x}_t$ along with their covariance matrix, $\mathbf P_t$ and the cross-covariance matrix $\mathbf M_t$ using \cite{Shumway2000}

@@

One second of data was generated using equations \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation} in response to an initial condition where $\boldsymbol\Sigma_v=0.1 \times \mathbf{I}_{n_y}$ and $ \eta(r-r')=1\times\phi_{4,3,-2}(r-r')$. Sampling time and the synaptic time constant are set to 1 and 10 milliseconds respectively. The sensors were arranged regularly where the distance between adjacent sensors was set to 0.5 mm resulting into 161 observation locations. Sensors are modelled using $\phi_{4,4,k}$, providing a width of 0.8 mm at half the maximum amplitude of the observation kernel, allowing for some cross talk in the recordings. The connectivity kernel and its decomposition are plotted in  \figurename{\ref{fig:KernelAndFreqResponse}} (a). The spatial frequency of the observed field is shown in \figurename{\ref{fig:KernelAndFreqResponse}} (b). The observed field has an approximate spatial bandwidth $1.38$ cycles/mm. This suggests the use of wavelets up to level $j=3$. Field reconstruction at different level of approximations is shown in \figurename{\ref{fig:FieldEstimates}}. The figure confirms that $j=3$ is adequate to capture the dynamic of the underlying field. Root mean square error (RMSE) of the field estimates is shown in \figurename{\ref{fig:RMSE}}. The histogram of the parameter estimates in \figurename{\ref{fig:ParametersHistogram}} is generated using 100 runs of the EM algorithm, although the estimates of $\theta_1$ is biased (7.4 \%) the RMSE between the true and estimated kernels is $1.83$.

% is obtained  and The dynamics associated with the IDE defined by (\ref{eq:1DSimulationKernel}) will be oscillatory and dissipative to initial conditions. The field is decomposed using scaling functions $\phi_{4;0,k} $ and the associated wavelet functions $\psi_{4;j,k} $ with $j=0,1$.
\begin{figure}[!h]
 \centering
 \includegraphics[scale=1]{./Graph/ObservationFreqResponse.pdf}
 \caption{(a) Mexican-hat connectivity kernel. The kernel is a composite of two B-spline
functions (dashed lines), representing local excitation and lateral inhibition. (b) The average (over time) power in dB of the spatial frequency of the observations generated by the kernel in (a). The dashed line shows the cutoff frequency (-3 dB point).}
\label{fig:KernelAndFreqResponse} 
  \end{figure} 
\begin{figure}[!h] 
 \centering
 %\includegraphics[width=0.5\textwidth]{./Graph/RMSE.pdf}
\includegraphics[scale=1]{./Graph/RMSE.pdf}
 \caption{RMSE of the estimated field at different spatial resolution.}
 \label{fig:RMSE}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Field.pdf}
 \caption{Spatial field at one time instant for different spatial resolutions, estimated and true fields are
shown by dashed and solid lines respectively; (a) $j=0$, $n_x=17$. (b) $j=1$, $n_x=33$. (c) $j=2$, $n_x=65$. (d) $j=3$, $n_x=131$. (e) $j=4$, $n_x=263$.}
 \label{fig:FieldEstimates}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Hist.pdf}
 \caption{Histograms of the parameter estimates over 100 realizations. In each of the subplots
the solid lines show the actual parameter values and the dotted lines show the means of the estimated
parameters. Left panel: The histogram of the central excitatory connectivity kernel basis function parameter
estimates, $\theta_0$. Right panel: The histogram of the surround inhibition connectivity kernel basis function parameter
estimates, $\theta_1$.}
 \label{fig:ParametersHistogram}
 \end{figure}

>>>>

@@

The  $m$th order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 
\begin{eqnarray}
N_{m}\left(s\right)&=&\left(N_{m-1}\ast N_{1}\right)\left(s\right)\nonumber \\
&=&\int_0^{1} N_{m-1}\left( s-r\right)dr \quad m>1 
\label{SplineConvolutionIntegral}
\end{eqnarray}
Here $\ast$ denotes the convolution operator and $N_1\left(s\right)$ is the characteristic function of the unit interval  $\left[ 0,1\right) $
\begin{equation}
N_{1}\left(s\right)=
\begin{cases}
1 & \text{if $ 0\le s<1$}, \\
0 & \mathrm{elsewhere}
\end{cases}
\end{equation}
In order to compute the  integrals (\ref{eq:Lambdax}) and (\ref{eq:Lambdatheta}) we need to be able to calculate the convolution and inner product of B-spline scaling functions. Equation (\ref{SplineConvolutionIntegral}) can be rewritten as $(m-1)$ times convolution of indicator function with itself 
\begin{equation}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right)
\end{equation}
then by using the associativity property of the convolution operator, it is clear that
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right)
\end{align}
A direct consequence of \eqref{eq:BsplineConvolution} is \cite{Chui1992} 
\begin{eqnarray}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle&=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
&=&N_{m+m'}\left(m'+l_{2}-l_{1}\right)
\label{eq:BsplineInnerProduct}
\end{eqnarray}
where $\left\langle  .,.\right\rangle $ denotes the inner product.

@@

\section{State and Parameter Estimation}
A state-space representation of the IDE model allows the use of the Expectation Maximisation (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the kernel and the field from the electrophysiological data. The EM algorithm is a two part iterative algorithm that yields the maximum likelihood estimate of the connectivity kernel parameters and the posterior distribution of the field weights. In the M-step a lower bound $\mathcal{Q}\left(\boldsymbol\theta,\boldsymbol\theta'\right)$ of the complete-data log-likelihood  $p\left(\mathbf{X},\mathbf{Y} \mid \boldsymbol\theta \right)$, is maximised under the assumption of the system parameters being $\boldsymbol\theta'$. In the E-step, given an estimate of $\boldsymbol\theta$, the Rauch Tung Striebel (RTS) smoother \cite{Gibsona2005} can be employed to determine the posterior state sequence. 

@@

where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The second terms in equations (\ref{eq:upsilon0}-\ref{eq:Upsilon}) can be computed before EM implementation which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}. $\Xi$-variables can be calculated using the smoother outputs, i.e., state estimates, $\hat{\mathbf x}_t$ along with their covariance matrix, $\mathbf P_t$ and the cross-covariance matrix $\mathbf M_t$ using \cite{Gibsona2005}

@@

One second of data was generated using equations \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation} in response to an initial condition where $\boldsymbol\Sigma_v=0.1 \times \mathbf{I}_{n_y}$ and $ \eta(r-r')=1\times\phi_{4,3,-2}(r-r')$. Sampling time and the synaptic time constant are set to 1 and 10 milliseconds respectively. The sensors were arranged regularly where the distance between adjacent sensors was set to 0.5 mm resulting into 161 observation locations. Sensors are modelled using $\phi_{4,4,k}$, providing a width of 0.8 mm at half the maximum amplitude of the observation kernel, allowing for some cross talk in the recordings. The connectivity kernel and its decomposition are plotted in  \figurename{\ref{fig:KernelAndFreqResponse}} (a). The spatial frequency of the observed field is shown in \figurename{\ref{fig:KernelAndFreqResponse}} (b). The observed field has an approximate spatial bandwidth $1.38$ cycles/mm. This suggests the use of wavelets up to level $j=3$. Field reconstruction at different level of approximations is shown in \figurename{\ref{fig:FieldEstimates}}. The figure confirms that $j=3$ is adequate to capture the dynamic of the underlying field. Root mean square error (RMSE) of the field estimates is shown in \figurename{\ref{fig:RMSE}}. The histogram of the parameter estimates in \figurename{\ref{fig:ParametersHistogram}} is generated using 100 runs of the EM algorithm, although the estimates of $\theta_1$ is biased (7.4 \%) the RMSE between the true and estimated kernels is $1.83$.
\begin{figure}[!h]
 \centering
 \includegraphics[scale=1]{./Graph/ObservationFreqResponse.pdf}
 \caption{(a) Mexican-hat connectivity kernel. The kernel is a composite of two B-spline
functions (dashed lines), representing local excitation and lateral inhibition. (b) The average (over time) power in dB of the spatial frequency of the observations generated by the kernel in (a). The dashed line shows the cutoff frequency (-3 dB point).}
\label{fig:KernelAndFreqResponse} 
  \end{figure} 
\begin{figure}[!h] 
 \centering
 %\includegraphics[width=0.5\textwidth]{./Graph/RMSE.pdf}
\includegraphics[scale=1]{./Graph/RMSE.pdf}
 \caption{RMSE of the estimated field at different spatial resolution.}
 \label{fig:RMSE}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Field.pdf}
 \caption{Spatial field at one time instant for different spatial resolutions, estimated and true fields are
shown by dashed and solid lines respectively; (a) $j=0$, $n_x=17$. (b) $j=1$, $n_x=33$. (c) $j=2$, $n_x=65$. (d) $j=3$, $n_x=131$. (e) $j=4$, $n_x=263$.}
 \label{fig:FieldEstimates}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Hist.pdf}
 \caption{Histograms of the parameter estimates over 100 realizations. In each of the subplots
the solid lines show the actual parameter values and the dotted lines show the means of the estimated
parameters. Left panel: The histogram of the central excitatory connectivity kernel basis function parameter
estimates, $\theta_0$. Right panel: The histogram of the surround inhibition connectivity kernel basis function parameter
estimates, $\theta_1$.}
 \label{fig:ParametersHistogram}
 \end{figure}
